Object-based Auditory Scenes Similarity Retrieval and Classification With Wavelet Scattering

### 2016-07-12 paper/structureScene.tex
Edit scattering section

Scattering transforms are time-shift invariant representations of audio signals consisting of auditory and modulation filter banks interspersed with complex modulus nonlinearities.
This section highlights the importance of invariance to time shifts and stability to time warps in the representation of acoustic scenes, and explains how the scattering transform is designed to satisfy these properties while having a high discriminative power.

\subsection{Invariance and stability in acoustic scenes}
The notion of invariance to time shifts plays an essential role in acoustic scene similarity retrieval as well as acoustic scene classification.
Indeed, recordings may be shifted locally without affecting their perception and therefore such shifts do not convey any information about the class.
To discard this superfluous source of variability, signals can be mapped to a time-shift invariant feature space before training the classifier, eliminating the need for the classifier to explicitly learn this invariance.
From any set of descriptors $\boldsymbol{x_1}(t,\gamma)$, where $\gamma$ denotes a descriptor index, a representation invariant to time shifts shorter in duration than $T$ can be obtained by convolving $\boldsymbol{x_1}$ with a low-pass filter $\boldsymbol{\phi}(t)$ of cutoff frequency set to $1/T$, as measured in Hertz:
\begin{equation}
\mathbf{S_1}\boldsymbol{x}(t, \gamma) = (\boldsymbol{x_1} \ast \boldsymbol{\phi}) (t).
\end{equation}
A downside is that the transient information in $\boldsymbol{x_1}$ at scales finer than $T$ are lost by this low-pass filtering, reducing discriminability in feature space.
To address this issue, the scattering transform recovers this information by convolving $\boldsymbol{x_1}$ with wavelets whose center frequencies are above $1/T$ and then applying complex modulus.

@@

Starting from a monophonic waveform $\boldsymbol{x}(t)$, the scattering transform is defined as an infinite cascade of wavelet transform and modulus operators.
However, to achieve invariance to translation up to $T = 372\,\mathrm{ms}$, \ie the approximately minimal duration between non-overlapping acoustic events, two layers of scattering transform.

@@

Let $\boldsymbol{\psi}(t)$ a complex-valued band-pass filter of
center frequency $\xi_1$ and bandwidth $\xi_1/Q_1$.
A filter bank of wavelets is built by dilating $\boldsymbol{\psi}(t)$
according to a geometric sequence of scales $2^{\gamma_1/Q_1}$, obtaining
\begin{equation}
\boldsymbol{\psi_{\gamma_1}}(t) = 2^{-\gamma_1/Q_1} \boldsymbol{\psi}(2^{-\gamma_1/Q_1} t)\mbox{.}
\end{equation}
The variable $\gamma_1$ is the opposite number of a log-frequency, taking integer values between $0$ and $(J_1 \times Q_1 - 1)$.
In the sequel, we set $\xi_1$ to $20~\mathrm{kHz}$ (close to the Nyquist frequency of the audio recordings) the number of octaves $J_1$ to $10$ (the lower end of human hearing range) and the number of wavelets per octave $Q_1$ to $8$.
For each $\gamma_1$, the wavelet $\boldsymbol{\psi_{\gamma_1}}(t)$
has a center frequency of $2^{-\gamma_1/Q_1}\xi_1$, a bandwidth of $2^{-\gamma_1/Q_1}\xi_1/Q_1$, and a quality factor of $Q_1$.

The wavelet transform of an audio signal
$\boldsymbol{x}(t)$ is obtained by convolution with all wavelets.
Applying pointwise complex modulus the transform yields
the wavelet scalogram
\begin{equation}
\boldsymbol{x_1}(t, \gamma_1)
= \vert \boldsymbol{x} \ast \boldsymbol{\psi_{\gamma_1}} \vert (t)\mbox{.}
\end{equation}
The wavelet scalogram bears resemblance to the constant-Q transform (CQT),
which is derived from the short-term Fourier transform (STFT) by averaging the frequency
axis into constant-Q subbands of center frequencies $2^{-\gamma_1/Q_1}\xi_1$.
Indeed, both time-frequency representations are indexed by time $t$ and log-frequency $\gamma_1$.
However, contrary to the CQT, the wavelet scalogram reaches the Heisenberg
theoretical limit of optimal time-frequency localization across the whole
frequency range, whereas the temporal resolution of the CQT, at least in its original implementation \cite{Brown1992}, is fixed by the support of the STFT analyzing window.
Therefore, the wavelet scalogram has a better temporal localization at high
frequencies than the CQT, at the expense of a greater computational cost.
This increase in time-frequency localization comes at the expense of a greater computational cost, the inverse fast Fourier transform (IFFT) routine must be called for each wavelet $\boldsymbol{\psi_{\gamma_1}}$ in the filter bank.
Yet, it allows us to observe amplitude modulations at fine temporal scales in the scalogram, down to the minimum scale $2Q_1/\xi_1$ for $\gamma_1 = 0$, of the order of $1\,\textrm{ms}$ given the aforementioned values of $Q_1$ and $\xi_1$.

\subsection{Extracting modulations with second-order scattering}
Among auditory scenes, amplitude modulations may be caused by a broad variety of mechanical interactions, including collision, friction, and turbulent flow.
At a longer scale, they also account for some higher-level attributes of sound, \eg prosody in speech or rhythm in music.
Although they are discarded while filtering $\boldsymbol{x_1}(t,\gamma_1)$ into a time-shift invariant representation $\mathbf{S_1}\boldsymbol{x}(t,\gamma_1)$, they can be recovered by a second wavelet transform modulus operator.
The amplitude modulation spectrum resulting from this operator is
\begin{equation}
\boldsymbol{x_2}(t,\gamma_1,\gamma_2) =
\vert \boldsymbol{x_1} \ast \boldsymbol{\psi_{\gamma_2}} \vert(t,\gamma_1),
\end{equation}
where the center frequencies of the wavelets $\boldsymbol{\psi_{\gamma_2}}(t)$ are of the form $2^{-\gamma_2/Q_2} \xi_2$, and the second-order log-frequency index $\gamma_2$ takes integer values between $0$ and $(J_2 \times Q_2 - 1)$. Note that these second-order wavelets are dilated versions of a second mother wavelet $\boldsymbol{\psi}$, with a different center frequency $\xi_2$ and quality factor $Q_2$. The identity of the wavelet will be clear from context.
In the sequel, we set $\xi_2$ to $2.5\,\mathrm{kHz}$, $Q_2$ to $1$, and $J_2$ to $12$. Lastly, the low-pass filter $\phi(t)$ is applied to $\boldsymbol{x_2}$ to guarantee invariance to time-shifting, which yields
\begin{equation}
\mathbf{S_2}\boldsymbol{x}(t,\gamma_1,\gamma_2) =
(\boldsymbol{x_2} \ast \phi)(t,\gamma_1,\gamma_2).
\end{equation}
The scattering transform $\mathbf{S}\boldsymbol{x}(t,\gamma)$ consists of the concatenation of first-order coefficients $\mathbf{S_1}\boldsymbol{x}(t,\gamma_1)$ and second-order coefficients $\mathbf{S_1}\boldsymbol{x}(t,\gamma_1,\gamma_2)$ into a feature matrix $\mathbf{S}\boldsymbol{x}(t,\gamma)$, where $\gamma$ is a shorthand for either $\gamma_1$ or $(\gamma_1,\gamma_2)$.

\subsection{Gammatone wavelets}
\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{gammatones}
\caption{
\label{fig:gammatones}
Gammatone wavelets $\psi(t)$ in the time domain with quality factors (a) $Q = 4$ and (b) $Q = 1$. Blue and red oscillations represent the real and imaginary parts. The orange envelope represents the complex modulus.}
\end{center}
\end{figure}
Wavelets
$\boldsymbol{\psi_{\gamma_1}}(t)$ and $\boldsymbol{\psi_{\gamma_2}}(t)$ are designed as fourth-order Gammatone
wavelets with one vanishing moment \cite{Venkitaraman2014}, as shown in Figure \ref{fig:gammatones}.
In the context of auditory scene analysis, the asymmetric envelope of Gammatone wavelets is more biologically plausible than the symmetric, Gaussian-like envelope of the more widely used Morlet wavelets.
Indeed, it allows to reproduce two important psychoacoustic effects in the mammalian cochlea: the asymmetry of temporal masking and the asymmetry of spectral masking  \cite{Fastl2007}.
Moreover, it should be noted that Gammatone wavelets follow the typical amplitude profile of natural sounds, beginning with a relatively sharp attack and ending with a slower decay.
As such, they can be discovered automatically by unsupervised encoding of natural sounds \cite{Smith2006}.
This suggests that, despite being hand-crafted and not learned, Gammatone wavelets provide a sparse time-frequency representation of acoustic scenes.

>>>>

Scattering transforms are time-shift invariant representations of audio signals consisting of auditory and modulation filter banks interspersed with complex modulus nonlinearities.
This section highlights the importance of invariance to time-shifts and stability to time-warping in the representation of acoustic scenes, and explains how the scattering transform is designed to satisfy these properties while having a high discriminative power.

\subsection{Invariance and stability in acoustic scenes}
The notion of invariance to time shifts plays an essential role in acoustic scene similarity retrieval as well as acoustic scene classification.
Indeed, recordings may be shifted locally without affecting their perception and therefore such shifts do not convey any information about the class.
To discard this superfluous source of variability, signals can be mapped to a time-shift invariant feature space before training the classifier, eliminating the need for this classifier to explicitly learn this invariance.
From any time-varying set of descriptors $\boldsymbol{x_1}(t,\gamma)$, where $\gamma$ denotes a descriptor index, a representation invariant to time-shifts shorter in duration than $T$ can be obtained by convolving $\boldsymbol{x_1}$ with a low-pass filter $\boldsymbol{\phi}(t)$ of cutoff frequency set to $1/T$, as measured in Hertz:
\begin{equation}
\mathbf{S_1}\boldsymbol{x}(t, \gamma) = (\boldsymbol{x_1} \ast \boldsymbol{\phi}) (t).
\end{equation}
A downside is that the transient information in $\boldsymbol{x_1}$ at scales finer than $T$ are lost by this low-pass filtering, reducing discriminability in feature space.
To address this issue, the scattering transform recovers this information by convolving $\boldsymbol{x_1}$ with wavelets whose center frequencies are above $1/T$ and then applying complex modulus.

@@

Starting from a monophonic waveform $\boldsymbol{x}(t)$, the scattering transform is defined as an infinite cascade of wavelet transform and modulus operators.
However, to achieve invariance to translation up to $T = 372\,\mathrm{ms}$, \ie the approximate minimal duration between non-overlapping acoustic events, two layers of scattering transform often suffice.

@@

Let $\boldsymbol{\psi}(t)$ a complex-valued band-pass filter of
center frequency $\xi_1$ and bandwidth $\xi_1/Q_1$.
A filter bank of wavelets is built by dilating $\boldsymbol{\psi}(t)$
according to a geometric sequence of scales $2^{\gamma_1/Q_1}$, obtaining
\begin{equation}
\boldsymbol{\psi_{\gamma_1}}(t) = 2^{-\gamma_1/Q_1} \boldsymbol{\psi}(2^{-\gamma_1/Q_1} t)\mbox{.}
\end{equation}
The variable $\gamma_1$ is a scale, or an inverse log-frequency, taking integer values between $0$ and $(J_1 \times Q_1 - 1)$.
In the sequel, we set $\xi_1$ to $20~\mathrm{kHz}$ (close to the Nyquist frequency of the audio recordings), the number of octaves $J_1$ to $10$ (the lower end of human hearing range), and the number of wavelets per octave $Q_1$ to $8$.
For each $\gamma_1$, the wavelet $\boldsymbol{\psi_{\gamma_1}}(t)$
has a center frequency of $2^{-\gamma_1/Q_1}\xi_1$, a bandwidth of $2^{-\gamma_1/Q_1}\xi_1/Q_1$, and a quality factor of $Q_1$.

The wavelet transform of an audio signal
$\boldsymbol{x}(t)$ is obtained by convolution with all wavelets.
Applying pointwise complex modulus the transform yields
the wavelet scalogram
\begin{equation}
\boldsymbol{x_1}(t, \gamma_1)
= \vert \boldsymbol{x} \ast \boldsymbol{\psi_{\gamma_1}} \vert (t)\mbox{.}
\end{equation}
The wavelet scalogram bears resemblance to the constant-Q transform (CQT),
which is derived from the short-term Fourier transform (STFT) by averaging the frequency
axis into constant-Q subbands of center frequencies $2^{-\gamma_1/Q_1}\xi_1$.
Indeed, both time-frequency representations are indexed by time $t$ and log-frequency $\gamma_1$.
However, contrary to the CQT, the wavelet scalogram reaches the Heisenberg
theoretical limit of optimal time-frequency localization across the whole
frequency range, whereas the temporal resolution of the traditional CQT is fixed by the support of the STFT analyzing window \cite{Brown1992}.
Therefore, the wavelet scalogram has a better temporal localization at high
frequencies than the CQT, at the expense of a greater computational cost
since the inverse fast Fourier transform (IFFT) routine must be called for each wavelet $\boldsymbol{\psi_{\gamma_1}}$ in the filter bank.
However, this allows us to observe amplitude modulations at fine temporal scales in the scalogram, down to the minimum scale $2Q_1/\xi_1$ for $\gamma_1 = 0$, of the order of $1\,\textrm{ms}$ given the aforementioned values of $Q_1$ and $\xi_1$.

\subsection{Extracting modulations with second-order scattering}
Among auditory scenes, amplitude modulations may be caused by a broad variety of mechanical interactions, including collision, friction, and turbulent flow.
At longer scales, they also account for higher-level attributes of sound, such as prosody in speech or rhythm in music.
Although they are discarded while filtering $\boldsymbol{x_1}(t,\gamma_1)$ into a time-shift invariant representation $\mathbf{S_1}\boldsymbol{x}(t,\gamma_1)$, they can be recovered by a second wavelet transform modulus operator.
The amplitude modulation spectrum resulting from this operator is
\begin{equation}
\boldsymbol{x_2}(t,\gamma_1,\gamma_2) =
\vert \boldsymbol{x_1} \ast \boldsymbol{\psi_{\gamma_2}} \vert(t,\gamma_1),
\end{equation}
where the center frequencies of the wavelets $\boldsymbol{\psi_{\gamma_2}}(t)$ are of the form $2^{-\gamma_2/Q_2} \xi_2$, and the second-order scale index $\gamma_2$ takes integer values between $0$ and $(J_2 \times Q_2 - 1)$. Note that these second-order wavelets are dilated versions of a second mother wavelet $\boldsymbol{\psi}$, with a different center frequency $\xi_2$ and quality factor $Q_2$. The identity of the wavelet will be clear from context.
In the sequel, we set $\xi_2$ to $2.5\,\mathrm{kHz}$, $Q_2$ to $1$, and $J_2$ to $12$. Lastly, the low-pass filter $\phi(t)$ is applied to $\boldsymbol{x_2}$ to guarantee invariance to time-shifting, which yields
\begin{equation}
\mathbf{S_2}\boldsymbol{x}(t,\gamma_1,\gamma_2) =
(\boldsymbol{x_2} \ast \phi)(t,\gamma_1,\gamma_2).
\end{equation}
The scattering transform $\mathbf{S}\boldsymbol{x}(t,\gamma)$ consists of the concatenation of first-order coefficients $\mathbf{S_1}\boldsymbol{x}(t,\gamma_1)$ and second-order coefficients $\mathbf{S_1}\boldsymbol{x}(t,\gamma_1,\gamma_2)$ into a feature matrix $\mathbf{S}\boldsymbol{x}(t,\gamma)$, where $\gamma$ is a shorthand for either $\gamma_1$ or $(\gamma_1,\gamma_2)$.

\subsection{Gammatone wavelets}
\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{gammatones}
\caption{
\label{fig:gammatones}
Gammatone wavelets $\psi(t)$ in the time domain with quality factors (a) $Q = 4$ and (b) $Q = 1$. Blue and red oscillations represent the real and imaginary parts. The orange envelope represents the complex modulus.}
\end{center}
\end{figure}
Wavelets
$\boldsymbol{\psi_{\gamma_1}}(t)$ and $\boldsymbol{\psi_{\gamma_2}}(t)$ are designed as fourth-order Gammatone
wavelets with one vanishing moment \cite{Venkitaraman2014}, and are shown in Figure \ref{fig:gammatones}.
In the context of auditory scene analysis, the asymmetric envelope of Gammatone wavelets is more biologically plausible than the symmetric, Gaussian-like envelope of the more widely used Morlet wavelets.
Indeed, it allows to reproduce two important psychoacoustic effects in the mammalian cochlea: the asymmetry of temporal masking and the asymmetry of spectral masking  \cite{Fastl2007}.
Moreover, it should be noted that Gammatone wavelets follow the typical amplitude profile of natural sounds, beginning with a relatively sharp attack and ending with a slower decay.
As such, they can be discovered automatically by unsupervised encoding of natural sounds \cite{Smith2006}.
This suggests that, despite being hand-crafted and not learned, Gammatone wavelets provide a sparse time-frequency representation of acoustic scenes.
